Elaboration with First-Class Implicit Function Types

### 2020-06-05 paper.tex
revision

\subsection{Contributions}
\begin{itemize}
  \item We propose an elaboration algorithm which translates from a small
    Agda-like surface language to a small Martin-L\"of type theory extended with
    implicit function types, telescopes and \emph{strictly curried function
      types} with telescope domain. We use these extensions to accumulate
    information about implicit insertions. Our algorithm is based on Norell's
    bidirectional elaborator for Agda \cite[Chapter~3]{norell07thesis}.
  \item
    In the System F fragment, the presented elaborator is comparable or superior
    to previous solutions for impredicative inference. However, it also supports
    full dependent type theory. Our inference is also global, i.e.\ it can
    consider the whole program and not just particular n-ary applications.
  \item We provide an executable implementation of the elaborator described in
    this paper.
  \item Our solution is simple: we implemented elaboration,
    evaluation and unification in about 670 lines of Haskell, of which 215 lines
    implement the novel enhancements on the top of Norell's basic elaborator.
  \item The target theory of elaboration serves as a general platform for
    elaborating implicit function types: our concrete elaborator is a relatively
    simple one, and there is room to further develop it.
\end{itemize}

@@

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. Our main goal with the elaborator is to have a setup
which is as simple as possible, but sufficient to illustrate the enhancements in
Section \ref{sec:extending_elaboration}, and make it clear that this could be
used to conservatively extend elaboration in existing dependently typed
languages. Hence, we do not have more fine-grained control over metavariable
insertion, nor the advanced unification algorithms used in Agda
\cite{abel2011higher}; all of these are orthogonal improvements to the presented
system.

@@

\begin{definition}[Fresh meta creation]\label{def:freshmeta} We define $\freshMeta{\Theta}{\Gamma}{A}$ as follows:
  \begin{mathpar}
  \inferrule*{\Theta|\Gamma\vdash A : \U}
             {\freshMeta{\Theta}{\Gamma}{A} \in \{(\Theta',\,\theta,\,t)\,|\,(\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t : A[\theta])\} \\
              \freshMeta{\Theta}{\Gamma}{A} :\equiv
              ((\Theta,\,\alpha:\Gamma\To A),\,\p,\,\alpha\,\overline{\mathsf{vars}_{\Gamma}})}
  \end{mathpar}
In the definition of $\freshMeta{\Theta}{\Gamma}{A}$, we extend $\Theta$ with a
fresh meta, which has the closing type $\Gamma \To A$. The $\p$ weakening
relates the new metacontext to the old one, by ``dropping'' the new
entry. Lastly, $\alpha\,\overline{\mathsf{vars}_{\Gamma}}$ is the fresh meta
applied to all bound variables.
\end{definition}

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have implicit
argument insertion as a helper function, which is defined by recursion on core
types, and which is used as a post-processing step after inference. First, about
the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insertion,flushleft]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert'(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           \\
            \einsert(\Theta,\,\theta,\,t,\,A)\hspace{0.3em}\in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           }
\end{mathpar}
\begin{alignat*}{3}
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert'\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert'\,\fail && :\equiv && \fail\\
  &\einsert\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A) && :\equiv&& \ereturn\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A)\\
  &\einsert\,(\Theta,\,\theta,\,t,\,A) && :\equiv&& \einsert'\,(\Theta,\,\theta,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail\\
\end{alignat*}
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

@@

The difference between $\einsert$ and $\einsert'$ is that the latter avoids
insertion to implicit $\lambda$-terms. We usually want to avoid that, as it
creates a new implicit $\beta$-redex where the fresh metavariable is often
under-constrained. We only create such $\beta$-redex when the surface program
explicitly asks for it; we shall shortly see this in the discussion of the
$\einferblank$ case for explicit function application. With this, we largely
follow the insertion behavior of Agda.

@@

We revisit now the $polyList$ example from Section \ref{sec:introduction}. We
assume the following:
\[
List : \U \to \U \qquad nil : \{A : \U\} \to List\,A \qquad
cons : \{A : \U\} \to A \to List\,A \to List\,A
\]
In Figure \ref{fig:polylist}, we present a trace of checking $cons\,(\lambda\,x.\,x)\,nil$
at type $List\,(\{A : \U\}\to A \to A)$. We omit context and metacontext
parameters everywhere, and mark recursive calls by indentation. We also omit
some checking, inference, implicit insertion and unification calls which are not
essential for illustration.

\begin{figure}
\begin{alignat*}{3}
  & \scriptstyle{0 }\qquad\qquad && \echeckt{cons\,(\lambda\,x.\,x)\,\mi{nil}}{(\mi{List}\,(\{A : \U\}\to A \to A))}
     \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}} \\
  & \scriptstyle{2 }  && \qquad \einfert{\mi{cons}\,(\lambda\,x.\,x)}\\
  & \scriptstyle{3 }  && \qquad\quad \einfert{\mi{cons}}\\
  & \scriptstyle{4 }  && \qquad\quad = \mi{cons}\,\{\alpha_0\}\,:\,\alpha_0 \to \mi{List}\,\alpha_0 \to \mi{List}\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = \mi{cons}\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x) : \mi{List}\,(\alpha_1\to\alpha_1) \to \mi{List}\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{8 }  && \qquad \echeckt{\mi{nil}}{(\mi{List}\,(\alpha_1\to\alpha_1))}\\
  & \scriptstyle{9 }  && \qquad = \mi{nil}\,\{\alpha_1\to\alpha_1\}\\
  & \scriptstyle{10} && \quad = \mi{cons}\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x)\,(\mi{nil}\,\{\alpha_1\to\alpha_1\}): \mi{List}\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{11} && \quad \unify\,(\mi{List}\,(\{A : \U\}\to A \to A))\,(\mi{List}\,(\alpha_1\to\alpha_1))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{13} && \qquad = \fail
\end{alignat*}
\caption{$\mi{polyList}$ elaboration.}
\label{fig:polylist}
\end{figure}

@@

We shall utilize the extended core theory to implement smarter
elaboration. Recall from Section \ref{sec:issues} that the old elaborator makes
two kinds of unforced insertion choices:
\begin{enumerate}
\item\label{itm:lamins} $\echeck{t}{\Theta}{\Gamma}{A}$ does not insert an implicit $\lambda$ when $A$ is meta-headed.
\item\label{itm:appins} $\einsert\,(\Theta,\,\theta,\,t,\,A)$ does not insert an implicit application when $A$ is meta-headed.
\end{enumerate}
In the following, we shall only enhance $\lambda$-insertions
(\ref{itm:lamins}). This allows a simple implementation which only requires
minimal changes to unification, and which is already remarkably powerful. It
seems that enhancing implicit application insertions in (\ref{itm:appins})
requires extending unification; we discuss this in Section
\ref{sec:appinsert}. First, we modify closing types and contextualization to
take advantage of telescopes.

\begin{definition}[Closing types]
We use curried function types to close over record types in the scope. If a
bound variable does not have a record type, then we do as before\footnote{This
  implies that we close over meta-headed types using plain functions. In theory,
  this causes a higher-order version of the basic implicit insertion problem: we
  are uncertain about whether we should be uncertain about implicit
  insertions. So far, this higher-order insertion problem seems to be
  irrelevant in practice, in the prototype implementation.}. We prepend the
following clause to Definition
\ref{def:closingtype}:
\[
  ((\Gamma,\, x : \Rec\,A) \To B) :\equiv (\Gamma \To (\{x : \ol{A}\} \to B))
\]
\end{definition}
\begin{definition}[Contextualization]
We extend spine notation to applications of curried functions. For example, we
may have a spine $\ol{t} \equiv (\{x : \ol{A}\},\,\{y : \ol{B}\})$. We
accordingly revise Definition \ref{def:contextualization} for
$\overline{\mathsf{vars}_{\Gamma}}$ so that we use curried function application
for each record type in $\Gamma$.
\end{definition}

@@

\subsection{Unification For Strictly Curried Functions}
Although we omit most details of unification, we shall discuss it for curried
functions, as it is essential in the extended elaboration algorithm. We seek to
solve meta-telescopes in a way such that iterated implicit function types on the
two sides of a unification problem become definitionally equal. First we define
a helper function which counts leading implicit function types:
\begin{alignat*}{3}
  & \implArity\,(\{x : A\}\to B) && :\equiv 1 + \implArity\,B\\
  & \implArity\,A                && :\equiv 0
\end{alignat*}
Then, we extend unification. The most interesting case is when we unify a
curried function type with an implicit function type. In this case, if there are
``missing'' implicit arguments on the side of the curried function type (note
the $\implArity$ condition on the pattern), we refine the $A$ telescope domain
to an extended $(x_0 : A_0) \TCons A_1$ telescope. Since we invent a fresh $A_1$
domain for a curried function type, we need to add a constancy constraint for it
as well.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x_0 : A_0\}\to B')\,\,|\,\,\implArity\,B < \implArity\,B' + 1 :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A_1) = \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
& \quad (\Theta_2,\,\theta_2) \leftarrow
  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A_0[\theta_1])\TCons A_1)\\
& \quad \slet\,(\Theta_3,\,\theta_3) =
    \newConstancy{\Theta_2}{\Gamma[\theta_{12}],\,x_0 : A_0[\theta_{12}]}{x_1}{(A_1[\theta_2])}{(B[\theta_{12}][x\mapsto (x_0 :: x_1)])}\\
& \quad\unify_{\Theta_3|\Gamma[\theta_{123}],\,x_0 : A_0[\theta_{123}]}\,
         (\{x_1 : A_1[\theta_{23}]\}\to B[\theta_{123}][x\mapsto (x_0 :: x_1)])\,B'
\end{alignat*}
We have the symmetric $\unify_{\Theta_0|\Gamma}(\{x_0 : A_0\}\to B')\,(\{x :
\ol{A}\}\to B)$ case the same way as above.

Below, let us assume that $B'$ is not meta-headed, and it also does not match
the pattern above. Hence, either $B'$ is not an implicit function type, or the
arity condition does not hold, so we learn that the domain telescope should be
empty.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
\end{alignat*}
Again, we also have the symmetric case. Some examples for unification behavior:
\begin{itemize}
\item If $B$ is not meta-headed, $\unify\,(\{x : \ol{\alpha}\}\to B)\,B$ causes
  $\alpha$ to be solved as $\epsilon$, for the following reason: if $B$ is an
  implicit function type, the arity condition fails, otherwise the pattern match
  fails, so either way we fall through the first rule and solve $\alpha$ to
  $\epsilon$.
\item If $B$ is again not meta-headed, $\unify\,(\{x : \ol{\alpha}\}\to B)\,(\{x
  : A\}\to B)$ causes $\alpha$ to be solved as $A \TCons \epsilon$.
\end{itemize}

@@

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
&\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow
            \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}) \\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term (avoiding superfluous insertion, as in
Section \ref{sec:elaboration} when inferring type for $\lambda$-s) and wrap the
result in a strictly curried $\lambda$. We again need to create new constancy
constraint. This way, if $B$ does not depend on $x : \Rec\,A$, the strictly
curried $\lambda$ in the output immediately computes away, since $A$ is solved
to $\epsilon$.

@@

We provide an implementation of the elaborator from Section
\ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
a surface expression from standard input, and outputs the result of elaboration,
or optionally the type or the normal form of the result. It is implemented in
1106 lines of Haskell. Of this, 677 lines constitute the core syntax,
evaluation, unification and elaboration. Of these 669 lines, 462 lines implement
the basic elaborator of Section \ref{sec:bidirectional_elaboration} and 215
lines implement the extended elaborator.

>>>>

\subsection{Contributions}
\begin{itemize}
  \item We propose an elaboration algorithm which translates from a small
    Agda-like surface language to a small Martin-L\"of type theory extended with
    implicit function types, telescopes and \emph{strictly curried function
      types} with telescope domain. We use these extensions to accumulate
    information about implicit insertions. Our algorithm is based on Norell's
    bidirectional elaborator for Agda \cite[Chapter~3]{norell07thesis}.
  \item
    In the System F fragment, the presented elaborator is comparable or superior
    to previous solutions for impredicative inference. However, it also supports
    full dependent type theory. Our inference is also global, i.e.\ it can
    consider the whole program and not just particular n-ary applications.
  \item We provide an executable implementation of the elaborator described in
    this paper.
  \item Our solution is simple: we implemented elaboration, evaluation and
    unification in about 680 lines of Haskell, of which about 230 lines
    implement the novel enhancements on the top of Norell's basic elaborator.
  \item The target theory of elaboration serves as a general platform for
    elaborating implicit function types: our concrete elaborator is a relatively
    simple one, and there is room to further develop it.
\end{itemize}

@@

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. Our main goal with the elaborator is to have a setup
which is as simple as possible, but sufficient to illustrate the enhancements in
Section \ref{sec:extending_elaboration}, which could be used to conservatively
extend elaboration in existing dependently typed languages. Hence, we do not
have more fine-grained control over metavariable insertion, nor the advanced
unification algorithms used in Agda \cite{abel2011higher}; these are orthogonal
improvements to the presented system.

@@

\begin{definition}[Fresh meta creation]\label{def:freshmeta} We define $\freshMeta{\Theta}{\Gamma}{A}$ as follows:
  \begin{mathpar}
  \inferrule*{\Theta|\Gamma\vdash A : \U}
             {\freshMeta{\Theta}{\Gamma}{A} \in \{(\Theta',\,\theta,\,t)\,|\,(\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t : A[\theta])\} \\
              \freshMeta{\Theta}{\Gamma}{A} :\equiv
              ((\Theta,\,\alpha:\Gamma\To A),\,\p,\,\alpha\,\overline{\mathsf{vars}_{\Gamma}})}
  \end{mathpar}
In the definition above, we extend $\Theta$ with a fresh meta, which has the
closing type $\Gamma \To A$. The $\p$ weakening relates the new metacontext to
the old one, by ``dropping'' the new entry. Lastly,
$\alpha\,\overline{\mathsf{vars}_{\Gamma}}$ is the fresh meta applied to all
bound variables.
\end{definition}

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have helper functions
for implicit argument insertion, which can be used as a post-processing step
after inference. First, about the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification. Patterns are matched in top-down order.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insertion,flushleft]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert'(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           \\
            \einsert(\Theta,\,\theta,\,t,\,A)\hspace{0.3em}\in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           }
\end{mathpar}
\begin{alignat*}{3}
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert'\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert'\,\fail && :\equiv && \fail\\
  &\einsert\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A) && :\equiv&& \ereturn\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A)\\
  &\einsert\,(\Theta,\,\theta,\,t,\,A) && :\equiv&& \einsert'\,(\Theta,\,\theta,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail\\
\end{alignat*}
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

@@

The difference between $\einsert$ and $\einsert'$ is that the former avoids
insertion to implicit $\lambda$-terms. We usually want to avoid that, as it
creates a new implicit $\beta$-redex where the fresh metavariable is often
under-constrained. We only create such $\beta$-redex when the surface program
explicitly asks for it; we shall shortly see this in the discussion of the
$\einferblank$ case for explicit function application. With this, we largely
follow the insertion behavior of Agda.

@@

We revisit now the $polyList$ example from Section \ref{sec:introduction}. We
assume the following:
\[
List : \U \to \U \qquad nil : \{A : \U\} \to List\,A \qquad
cons : \{A : \U\} \to A \to List\,A \to List\,A
\]
Below, we present a trace of checking $cons\,(\lambda\,x.\,x)\,nil$
at type $List\,(\{A : \U\}\to A \to A)$. We omit context and metacontext
parameters everywhere, and mark recursive calls by indentation. We also omit
some checking, inference, implicit insertion and unification calls which are not
essential for illustration.

%% \begin{figure}
\begin{alignat*}{3}
  & \scriptstyle{0 }\qquad\qquad && \echeckt{cons\,(\lambda\,x.\,x)\,\mi{nil}}{(\mi{List}\,(\{A : \U\}\to A \to A))}
     \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}} \\
  & \scriptstyle{2 }  && \qquad \einfert{\mi{cons}\,(\lambda\,x.\,x)}\\
  & \scriptstyle{3 }  && \qquad\quad \einfert{\mi{cons}}\\
  & \scriptstyle{4 }  && \qquad\quad = \mi{cons}\,\{\alpha_0\}\,:\,\alpha_0 \to \mi{List}\,\alpha_0 \to \mi{List}\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = \mi{cons}\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x) : \mi{List}\,(\alpha_1\to\alpha_1) \to \mi{List}\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{8 }  && \qquad \echeckt{\mi{nil}}{(\mi{List}\,(\alpha_1\to\alpha_1))}\\
  & \scriptstyle{9 }  && \qquad = \mi{nil}\,\{\alpha_1\to\alpha_1\}\\
  & \scriptstyle{10} && \quad = \mi{cons}\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x)\,(\mi{nil}\,\{\alpha_1\to\alpha_1\}): \mi{List}\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{11} && \quad \unify\,(\mi{List}\,(\{A : \U\}\to A \to A))\,(\mi{List}\,(\alpha_1\to\alpha_1))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\alpha_1\to\alpha_1)\\
  & \scriptstyle{13} && \qquad = \fail
\end{alignat*}
%% \caption{$\mi{polyList}$ elaboration.}
%% \label{fig:polylist}
%% \end{figure}

@@

We shall utilize the extended core theory to implement smarter
elaboration. Recall from Section \ref{sec:issues} that the old elaborator makes
two kinds of unforced insertion choices:
\begin{enumerate}
\item\label{itm:lamins} $\echeck{t}{\Theta}{\Gamma}{A}$ does not insert an implicit $\lambda$ when $A$ is meta-headed.
\item\label{itm:appins} $\einsert\,(\Theta,\,\theta,\,t,\,A)$ does not insert an implicit application when $A$ is meta-headed.
\end{enumerate}
In the following, we shall only enhance $\lambda$-insertions
(\ref{itm:lamins}). This allows a simple implementation which only requires
small changes to unification, and which is already remarkably powerful. It
seems that enhancing implicit application insertions in (\ref{itm:appins})
requires extending unification; we discuss this in Section
\ref{sec:appinsert}. First, we modify closing types and contextualization to
take advantage of telescopes.

\begin{definition}[Closing types]
We use curried function types to close over record types in the scope. If a
bound variable does not have a record type, then we do as before\footnote{This
  implies that we close over meta-headed types using plain functions. In theory,
  this causes a higher-order version of the basic implicit insertion problem: we
  are uncertain about whether we should be uncertain about implicit
  insertions. This issue does not seem to come up in realistic programs in the prototype implementation; the first-order treatment of uncertain insertions seem to leave little remaining ``higher-order'' uncertainty on the table }. We prepend the
following clause to Definition
\ref{def:closingtype}:
\[
  ((\Gamma,\, x : \Rec\,A) \To B) :\equiv (\Gamma \To (\{x : \ol{A}\} \to B))
\]
\end{definition}
\begin{definition}[Contextualization]
We extend spine notation to applications of curried functions. For example, we
may have a spine $\ol{t} \equiv (\{x : \ol{A}\},\,\{y : \ol{B}\})$. We
accordingly revise Definition \ref{def:contextualization} for
$\overline{\mathsf{vars}_{\Gamma}}$ so that we use curried function application
for each record type in $\Gamma$.
\end{definition}

@@

\subsection{Unification For Strictly Curried Function Types}
\label{sec:curriedunif}
Although we omit most details of unification, we shall discuss it for curried
function types, as it is essential in the extended elaboration algorithm. We
seek to solve meta-telescopes in a way such that iterated implicit function
types on the two sides of a unification problem become definitionally
equal. First we define a helper function which counts leading implicit function
types:
\begin{alignat*}{3}
  & \implArity\,(\{x : A\}\to B) && :\equiv 1 + \implArity\,B\\
  & \implArity\,A                && :\equiv 0
\end{alignat*}
Then, we extend unification. The most interesting case is when we unify a
curried function type with an implicit function type. In this case, if there are
``missing'' implicit arguments on the side of the curried function type (note
the $\implArity$ condition on the pattern), we refine the $A$ telescope domain
to an extended $(x_0 : A_0) \TCons A_1$ telescope. Since we invent a fresh $A_1$
domain for a curried function type, we need to add a constancy constraint for it
as well.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x_0 : A_0\}\to B')\,\,|\,\,\implArity\,B < \implArity\,B' + 1 :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A_1) = \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
& \quad (\Theta_2,\,\theta_2) \leftarrow
  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A_0[\theta_1])\TCons A_1)\\
& \quad \slet\,(\Theta_3,\,\theta_3) =
    \newConstancy{\Theta_2}{\Gamma[\theta_{12}],\,x_0 : A_0[\theta_{12}]}{x_1}{(A_1[\theta_2])}{(B[\theta_{12}][x\mapsto (x_0 :: x_1)])}\\
& \quad\unify_{\Theta_3|\Gamma[\theta_{123}],\,x_0 : A_0[\theta_{123}]}\,
         (\{x_1 : A_1[\theta_{23}]\}\to B[\theta_{123}][x\mapsto (x_0 :: x_1)])\,B'
\end{alignat*}
We have the symmetric $\unify_{\Theta_0|\Gamma}(\{x_0 : A_0\}\to B')\,(\{x :
\ol{A}\}\to B)$ case the same way as above.

Below, let us assume that $B'$ is not meta-headed, and it also does not match
the pattern above. In this case we solve the domain telescope to be empty.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
\end{alignat*}
Again, we also have the symmetric case. Some examples for unification behavior:
\begin{itemize}
\item If $B$ is not meta-headed, $\unify\,(\{x : \ol{\alpha}\}\to B)\,B$ causes
  $\alpha$ to be solved as $\epsilon$, for the following reason: if $B$ is an
  implicit function type, the arity condition fails, otherwise the pattern match
  fails, so either way we fall through the first rule and solve $\alpha$ to
  $\epsilon$.
\item If $B$ is again not meta-headed, $\unify\,(\{x : \ol{\alpha}\}\to B)\,(\{x
  : A\}\to B)$ causes $\alpha$ to be solved as $(x : A) \TCons \epsilon$.
\end{itemize}
Note that this algorithm matches one curried function type at a time, and does
not recognize that multiple possible solutions exist e.g.\ in the case of
$\unify\,(\{x : \ol{\alpha}\}\{y : \ol{\beta}\}\to C)\,(\{x : A\}\{y : B\}\to
C)$, where we could solve $\alpha$ to $\epsilon$, or $(x : A)\TCons\epsilon$, or
$(x : A)\TCons(y : B)\TCons\epsilon$. We leave such more sophisticated n-ary
matching to future work.

@@

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
&\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow
            \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}) \\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term and wrap the result in a strictly curried
$\lambda$. We again need to create new constancy constraint. This way, if $B$
does not depend on $x : \Rec\,A$, the strictly curried $\lambda$ in the output
immediately computes away, since $A$ is solved to $\epsilon$.

@@

We provide an implementation of the elaborator from Section
\ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
a surface expression from standard input, and outputs the result of elaboration,
or optionally the type or the normal form of the result. It is implemented in
about 1100 lines of Haskell. Of this, 680 lines constitute the core syntax,
evaluation, unification and elaboration. Of these 680 lines, about 450 lines
implement the basic elaborator of Section \ref{sec:bidirectional_elaboration}
and 230 lines implement the extended elaborator.

Elaboration is implemented in the style of Coquand's algorithm
\cite{coquand1996algorithm}, where elaboration is interleaved with
normalization-by-evaluation. Hence, we do not perform any substitution
operations on core syntax, instead we evaluate core terms into the semantic
domain, and perform unification, strengthening and occurrence checking on
semantic values. There is also a quoting (or readback) operation which yields
normalized core terms from values.
